A Standardized Effect Size for Evaluating the Comparing the Strength of Phylogenetic Signal

### 2020-07-15 Manuscript/OldVsns/Manuscript-1-DCA.tex
pnas vsn

For many macroevolutionary analyses, it is often of interest to quantify
the degree to which phylogenetic signal is displayed in continuous
traits. Several analytical tools have been developed for this purpose
(e.g., Gittleman and Kot 1990; Abouheif 1999; Pagel 1999; Blomberg et
al. 2003; Klingenberg and Gidaszewski 2010; Adams 2014a), which differ
primarily in how they characterize the phylogenetic dependency of trait
variation among taxa. One commonly used statistical measure,
\emph{Kappa} (\(K\)), expresses the strength of phylogenetic signal as
the ratio of observed trait variation to the trait variation conditioned
on the phylogeny; scaled by what is expected under Brownian motion given
the phylogeny's size and shape (Blomberg et al. 2003; also Adams 2014a).
Another approach, Pagel's \(\lambda\) (Pagel 1999), uses maximum
likelihood to fit the data to the phylogeny under some model of
evolutionary change (typically Brownian motion). The inclusion of a
scaling parameter, \(\lambda\), transforms the lengths of the internal
branches of the phylogeny to improve the fit, and this parameter
describes the degree of phylogenetic signal in the dataset (Pagel 1999;
Freckleton et al. 2002). Pagel's \(\lambda\) also has the appeal that it
may be included when estimating the association of traits in a
phylogenetic context, meaning that one may account for the degree of
phylogenetic signal while conducting phylogenetic regression or ANOVA
(see Freckleton et al. 2002). \hfill\break

Several studies have investigated the statistical properties of methods
for estimating phylogenetic signal under various conditions (e.g.,
Munkemuller et al. 2012; Pavoine and Ricotta 2012; Diniz-Filho et al.
2012; Molina-Venegas and Rodriguez 2017; see also Revell et al. 2008;
Revell 2010). These have largely focused on the ability of methods to
detect the presence of phylogenetic signal (i.e., type I and type II
error rates) under complex models of evolutionary change, across a range
of phylogeny sizes, and with varying degrees of phylogenetic uncertainty
or unresolved topologies. In terms of parameter estimation, Revell
(2010) found that regression parameters were accurately estimated when
\(\lambda\) was included during phylogenetic regression, and Munkmuller
et al.~(2012) demonstrated that estimates of phylogenetic signal
obtained using various measures generally increased when input levels of
phylogenetic signal were stronger. However, the precision of those
estimates could not be determined, because the input levels of
phylogenetic signal were simulated via a scaling factor (\(w\)) that was
not directly comparable to the measures of phylogenetic signal being
compared (see Munkemuller et al. 2012). One study (Boettiger et al.
2012) found that estimates of Pagel's \(\lambda\) displayed less
variation when data were simulated on a large phylogeny (\(N=281\)) as
compared to a small one (\(N=13\)), concluding that insufficient data
(i.e., the number of species) was the underlying cause of the increased
variation across parameter estimates. However, this conclusion assumes
that the precision in estimating \(\lambda\) remains constant across its
range (\(\lambda = 0 \to 1\)); an assumption that to date, has not been
verified. Thus, despite widespread use of Pagel's (1999) \(\lambda\) in
macroevolutionary studies, at present, we still lack a general
understanding of the precision with which \(\lambda\) can estimate
levels of phylogenetic signal in phenotypic datasets. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) in
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We also
evaluate the precision of a standardized effect size (\(Z_K\)) based on
\(Kappa\) for comparison. Additionally, we use simulations to determine
how the inclusion of \(\lambda\) in phylogenetic regression and ANOVA
(i.e., PGLS) affects parameter estimation, and whether levels of
phylogenetic signal estimated in a PGLS framework are accurate. We then
survey the recent macroevolutionary literature for published papers
containing estimates of \(\lambda\) from empirical datasets, and compare
these empirical estimates to patterns gleaned from our computer
simulations. In general we find that while PGLS parameters (e.g.,
\(\beta\)) are accurately estimated with the inclusion of phylogenetic
signal, estimates of \(\lambda\) are not. We also find that estimates of
\(\lambda\) vary widely for a given input value of phylogenetic signal,
and that this variation is not constant across the range of input
signal, with decreased precision when phylogenetic signal is of
intermediate strength. In addition, the same \(\lambda_{est}\) may be
obtained from datasets containing vastly different input levels of
phylogenetic signal. By contrast, variation across effect sizes
(\(Z_K\)) obtained from \(Kappa\) is far more consistent across the
range of input values, and the precision of \(Z_K\) is generally
preferable to that of \(\lambda_{est}\). Further, because \(Z_K\) is a
standardized effect size, the strength of phylogenetic signal is more
readily interpretable, and may be compared across datasets. We conclude
that estimates of phylogenetic signal using Pagel's \(\lambda\) are
often innacurate, and thus interpreting strength of phylogenetic signal
in phenotypic datasets based on this measure is compromised. By
contrast, effect sizes obtained from \(Kappa\) hold promise for
characterizing phylogenetic signal, and for comparing the strength of
phylogenetic signal across datasets.

>>>>

For many macroevolutionary analyses, it is often of interest to quantify
the degree to which phylogenetic signal is displayed in continuous
traits. Several analytical tools have been developed for this purpose
(e.g., Gittleman and Kot 1990; Abouheif 1999; Pagel 1999; Blomberg et
al. 2003; Klingenberg and Gidaszewski 2010; Adams 2014a), which differ
primarily in how they characterize the phylogenetic dependency of trait
variation among taxa. One commonly used statistical measure,
\emph{Kappa} (\(K\)), expresses the strength of phylogenetic signal as
the ratio of observed trait variation to the trait variation conditioned
on the phylogeny; which is then scaled by what is expected under
Brownian motion given the phylogeny's size and shape (Blomberg et al.
2003; also Adams 2014a). Another approach, Pagel's \(\lambda\) (Pagel
1999), uses maximum likelihood to fit the data to the phylogeny under
some model of evolutionary change (typically Brownian motion). The
inclusion of a scaling parameter, \(\lambda\), transforms the lengths of
the internal branches of the phylogeny to improve the fit, and this
parameter describes the degree of phylogenetic signal in the dataset
(Pagel 1999; Freckleton et al. 2002). Pagel's \(\lambda\) also has the
appeal that it may be included when estimating the association of traits
in a phylogenetic context, meaning that one may account for the degree
of phylogenetic signal while conducting phylogenetic regression or ANOVA
(see Freckleton et al. 2002). \hfill\break

Several studies have investigated the statistical properties of methods
for estimating phylogenetic signal under various conditions (e.g.,
Munkemuller et al. 2012; Pavoine and Ricotta 2012; Diniz-Filho et al.
2012; Molina-Venegas and Rodriguez 2017; see also Revell et al. 2008;
Revell 2010). These have largely focused on the ability of methods to
detect the presence of phylogenetic signal (i.e., type I and type II
error rates) under complex models of evolutionary change, across a range
of phylogeny sizes, and with varying degrees of phylogenetic uncertainty
or unresolved topologies. In terms of parameter estimation, Revell
(2010) found that regression parameters were accurately estimated when
\(\lambda\) was included during phylogenetic regression, and Munkmuller
et al.~(2012) demonstrated that estimates of phylogenetic signal
obtained using various measures generally increased when input levels of
phylogenetic signal were stronger. However, the precision of those
estimates could not be determined, because the input levels of
phylogenetic signal were simulated via a scaling factor (\(w\)) that was
not directly comparable to the measures of phylogenetic signal being
compared (see Munkemuller et al. 2012). Indeed, understanding the
precision in estimating levels of phylogenetic signal is paramount,
because biological inferences regarding the strength of phylogenetic
signal in phenotypic datasets requires high precision of its estimation.
One study (Boettiger et al. 2012) found that estimates of Pagel's
\(\lambda\) displayed less variation when data were simulated on a large
phylogeny (\(N=281\)) as compared to a small one (\(N=13\)), concluding
that insufficient data (i.e., the number of species) was the underlying
cause of the increased variation across parameter estimates. However,
this conclusion assumes that the precision in estimating \(\lambda\)
remains constant across its range (\(\lambda = 0 \to 1\)); an assumption
that to date, has not been verified. Thus, despite widespread use of
Pagel's (1999) \(\lambda\) in macroevolutionary studies, at present, we
still lack a general understanding of the precision with which
\(\lambda\) can estimate levels of phylogenetic signal in phenotypic
datasets. \hfill\break

In this study, we evaluate the precision of Pagel's \(\lambda\) in
estimating known levels of phylogenetic signal in phenotypic data. We
use computer simulations with differing numbers of species, differently
shaped phylogenies, and differing input levels of phylogenetic signal,
to explore the degree to which \(\lambda\) correctly identifies known
levels of phylogenetic signal, and under what circumstances. We also
evaluate the precision of a standardized effect size (\(Z_K\)) based on
\(Kappa\) for comparison. Additionally, we use simulations to determine
how the inclusion of \(\lambda\) in phylogenetic regression and ANOVA
(i.e., PGLS) affects parameter estimation, and whether levels of
phylogenetic signal estimated in a PGLS framework are accurate. We then
survey the recent macroevolutionary literature for published papers
containing estimates of \(\lambda\) from empirical datasets, and compare
these empirical estimates to patterns gleaned from our computer
simulations. In general we find that while PGLS parameters (e.g.,
\(\beta\)) are accurately estimated with the inclusion of phylogenetic
signal, estimates of \(\lambda\) are not. We also find that estimates of
\(\lambda\) vary widely for a given input value of phylogenetic signal,
and that this variation is not constant across the range of input
signal, with decreased precision when phylogenetic signal is of
intermediate strength. Additionally, the same \(\lambda_{est}\) may be
obtained from datasets containing vastly different input levels of
phylogenetic signal. By contrast, variation across effect sizes
(\(Z_K\)) obtained from \(Kappa\) is far more consistent across the
range of input values, and the precision of \(Z_K\) is generally
preferable to that of \(\lambda_{est}\). Further, because \(Z_K\) is a
standardized effect size, the strength of phylogenetic signal is more
readily interpretable, and may be compared across datasets. We conclude
that estimates of phylogenetic signal using Pagel's \(\lambda\) are
often inaccurate, and thus interpreting strength of phylogenetic signal
in phenotypic datasets based on this measure is compromised. By
contrast, effect sizes obtained from \(Kappa\) hold promise for
characterizing phylogenetic signal, and for comparing the strength of
phylogenetic signal across datasets.
